Simplifying Lotka--Volterra interaction matrices by species grouping: optimal groupings are unpredictable

### 2022-08-08 article.rnw
Set up xr to link from the article to the supplementary.
Reference the supplementary figures in the main text.
Update a bioRxiv article reference to the published version.

@@

One simplifying approach is to assume that most pairs of species do not interact or interact weakly, and hence setting many of the parameters to zero \parencite{mutshinda2009drives, weiss2021disentangling}.
However, weak interactions have been shown to be very important to the stability of ecological communities \parencite{mccann1998weak,neutel2002stability,wootton2016many}. Setting any individual parameter to zero thus risks losing important information; moreover, we often do not know with confidence which parameters are likely to be zeros and which are not.
Another approach assumes that species fall into groups, such that all species in the same group interact with species in other groups in an identical fashion, thereby collapsing pairwise interaction parameters for all the species in that group into sets of common parameters \parencite{ovaskainen2017species}.

In ecological contexts, such groups have been defined \emph{a priori} by phylogeny at some taxonomic level or by functional (i.e., trait) similarity \parencite{uriarte2004spatially,ovaskainen2017species,weiss2021disentangling,martyn2021identifying}.
The argument for grouping based on phylogeny relies on the principle that closely related species will fill similar ecological niches, and hence will interact with other species in a similar fashion \parencite{godoy2014phylogenetic,webb2002phylogenies,darwin1859origin}. %Darwin Chapter 3 "Struggle for Life Most Severe Between Individuals and Varieties of the Same Species".
However, empirical tests of this principle have found remarkably little support for it \parencite{godoy2014phylogenetic,cahill2008does,mayfield2010opposing}.
On the other hand, functional groupings have also proven to be poor predictors of species interactions \parencite{bret2008plant,funk2017revisiting}.
While attempts have been made to define functional groupings that are more useful in a modelling context \parencite{boulangeat2012improving}, these rely upon measurements of biotic interactions to define such groups.
As such, integrating these into species interaction models leaves us with a circular argument: a functional group is a group of species that interact with others in a similar way, and we therefore assume that all the species in a functional group interact with others in a similar way.
In all, we are left with the question of how best to group species in community models, and whether we can do so in a meaningful way using less data than required for a fully pairwise parameterisation.

@@

% DBS: Should we add a figure that shows the fit of observed vs predicted? Are we forgetting how much that might resonate with readers and jumping ahead too quickly?

% DBS: As per my comment immediately above, we need to add a reference to the partial residual plots in the supps.

The $R^2$ values of the models ranged from \rsquarednum{\Sexpr{min(brute_data$statistics$R2)}} to \rsquarednum{\Sexpr{max(brute_data$statistics$R2)}}.
The {\ic} values ranged from \icnum{\Sexpr{min(brute_data$statistics[[ic]])}} to \icnum{\Sexpr{max(brute_data$statistics[[ic]])}}.
The ungrouped model gave an {\ic} value of \icnum{\Sexpr{brute_data$fully_separated_statistics[[ic]]}}.
\Cref{fig:ic} shows the distribution of {\ic} values from the models, with some models of interest marked.

>>>>

\externaldocument{supp}

@@

One simplifying approach is to assume that most pairs of species do not interact or interact weakly, and hence setting many of the parameters to zero \parencite{mutshinda2009drives, weiss2022disentangling}.
However, weak interactions have been shown to be very important to the stability of ecological communities \parencite{mccann1998weak,neutel2002stability,wootton2016many}. Setting any individual parameter to zero thus risks losing important information; moreover, we often do not know with confidence which parameters are likely to be zeros and which are not.
Another approach assumes that species fall into groups, such that all species in the same group interact with species in other groups in an identical fashion, thereby collapsing pairwise interaction parameters for all the species in that group into sets of common parameters \parencite{ovaskainen2017species}.

In ecological contexts, such groups have been defined \emph{a priori} by phylogeny at some taxonomic level or by functional (i.e., trait) similarity \parencite{uriarte2004spatially,ovaskainen2017species,martyn2021identifying}.
The argument for grouping based on phylogeny relies on the principle that closely related species will fill similar ecological niches, and hence will interact with other species in a similar fashion \parencite{godoy2014phylogenetic,webb2002phylogenies,darwin1859origin}. %Darwin Chapter 3 "Struggle for Life Most Severe Between Individuals and Varieties of the Same Species".
However, empirical tests of this principle have found remarkably little support for it \parencite{godoy2014phylogenetic,cahill2008does,mayfield2010opposing}.
On the other hand, functional groupings have also proven to be poor predictors of species interactions \parencite{bret2008plant,funk2017revisiting}.
While attempts have been made to define functional groupings that are more useful in a modelling context \parencite{boulangeat2012improving}, these rely upon measurements of biotic interactions to define such groups.
As such, integrating these into species interaction models leaves us with a circular argument: a functional group is a group of species that interact with others in a similar way, and we therefore assume that all the species in a functional group interact with others in a similar way.
In all, we are left with the question of how best to group species in community models, and whether we can do so in a meaningful way using less data than required for a fully pairwise parameterisation.

@@

Plots of predicted against observed values and partial residual plots for the fitted alpha values are shown in \cref{fig:model-fit-best,fig:model-fit-average,fig:partial-residual-best,fig:partial-residual-average}.
The $R^2$ values of the models ranged from \rsquarednum{\Sexpr{min(brute_data$statistics$R2)}} to \rsquarednum{\Sexpr{max(brute_data$statistics$R2)}}.
The {\ic} values ranged from \icnum{\Sexpr{min(brute_data$statistics[[ic]])}} to \icnum{\Sexpr{max(brute_data$statistics[[ic]])}}.
The ungrouped model gave an {\ic} value of \icnum{\Sexpr{brute_data$fully_separated_statistics[[ic]]}}.
\Cref{fig:ic} shows the distribution of {\ic} values from the models, with some models of interest marked.
